---
title: "COLOC_Notes"
author: "Peter Fiorica"
date: "April 3, 2020"
output: html_document
---

```{r}
library(argparse)
library(data.table)
library(dplyr)
library(R.utils) #I forgot why I had this
"%&%" = function(a, b) paste(a, b, sep = "")
```


#Prepping GWAS Results for COLOC Use
```{r}
bim<-fread("/home/peter/prostate_cancer/liftover/hg38_imputed_aa_procan.bim", header = F)


GWAS_results<-fread("/home/peter/prostate_cancer/hg38_aa_procan_gwas.assoc.logistic", header = T)
frequency<-fread("/home/peter/prostate_cancer/liftover/hg38_imputed_aa_freq.frq", header = T)

GWAS_results$BETA<-log(GWAS_results$OR)
GWAS_results$SE<-((GWAS_results$BETA)/sqrt(abs(GWAS_results$STAT)))
GWAS_results$MAF<-frequency$MAF
GWAS_results$CHR<-paste("chr", GWAS_results$CHR, sep = "")
GWAS_results$SNP_ID<-paste(GWAS_results$CHR, GWAS_results$BP,bim$V6,bim$V5, "b38", sep = "_")

fwrite(GWAS_results, "/home/peter/prostate_cancer/colocalization/COLOC/AA_procan_GWASwBETASE.txt", col.names = T, row.names = F, sep = "\t", quote =F)
```

```
#Split file by chromosome
awk '{print $0 >> $1".txt"}' AA_procan_GWASwBETASE.txt
```

#COLOC Input Script

## Set up
```{r, eval = FALSE}
ma_prefix = ""
GWAS_prefix = ""
sample_size <- 4769
pheno_names <- "/home/peter/prostate_cancer/colocalization/COLOC/pheno_names.txt"

pheno_names <- fread(pheno_names, header = F)$V1
system("mkdir -p COLOC_input/")


#general setup data
chrs <- c(1:22)
tissues <- c("Adipose_Subcutaneous", "Adipose_Visceral_Omentum", "Adrenal_Gland", "Artery_Aorta", "Artery_Coronary", "Artery_Tibial", "Brain_Amygdala", "Brain_Anterior_cingulate_cortex_BA24", "Brain_Caudate_basal_ganglia", "Brain_Cerebellar_Hemisphere", "Brain_Cerebellum", "Brain_Cortex", "Brain_Frontal_Cortex_BA9", "Brain_Hippocampus", "Brain_Hypothalamus", "Brain_Nucleus_accumbens_basal_ganglia", "Brain_Putamen_basal_ganglia", "Brain_Spinal_cord_cervical_c-1", "Brain_Substantia_nigra", "Breast_Mammary_Tissue","Cells_Cultured_fibroblasts","Cells_EBV-transformed_lymphocytes" , "Colon_Sigmoid", "Colon_Transverse", "Esophagus_Gastroesophageal_Junction", "Esophagus_Mucosa", "Esophagus_Muscularis", "Heart_Atrial_Appendage", "Heart_Left_Ventricle", "Kidney_Cortex", "Liver", "Lung", "Minor_Salivary_Gland", "Muscle_Skeletal", "Nerve_Tibial", "Pancreas", "Pituitary", "Prostate", "Skin_Not_Sun_Exposed_Suprapubic", "Skin_Sun_Exposed_Lower_leg", "Small_Intestine_Terminal_Ileum", "Spleen", "Stomach", "Testis", "Thyroid", "Whole_Blood")             
tissues_sample_size <- c(581, 469, 233, 387, 213, 584, 129, 147, 194, 175, 209, 205, 175, 165, 170, 202, 170, 126, 114, 396, 483, 147, 318, 368, 330, 497, 465, 372, 386, 73, 208, 515, 144, 706, 532, 305, 237, 221, 517, 605, 174, 227, 324, 322, 574, 670)
GWAS_SNPs <- fread("/home/peter/prostate_cancer/colocalization/COLOC/chr_files/AA_procan_GWASwBETASE.txt", header = T)$SNP_ID


GWAS_write <- data.frame(panel_variant_id = character(), effect_size = numeric(), standard_error = numeric(), frequency = numeric(), sample_size = numeric(), stringsAsFactors = F) 
```
Above, we set up the general parameters for the COLOC input script. In the actual script, both the samples size and phenotype name file are specificied by flags.  They are explicitly written here for the sake of the example.  After those are specified, we make the output directory and specify chromosmes and tissues.  The tissues and sample sizes are all male tissues from GTEx V8.  We then read in a GWAS result file we made two chunks above.  After this, the next piece to the script does the actual bulk of it.

## Body of the Script
```{r, eval = FALSE}
for(tissue in 1:length(tissues)){ #read in tissue's .frq file for MAF
  print(tissue)
  eQTL_write <- data.frame(gene_id = character(), variant_id = character(), maf = numeric(), pval_nominal = numeric(), slope = numeric(), slope_se = numeric(), n_samples = numeric(), stringsAsFactors = F) #Creates a blank eQTL data table for a new tissue
  meqtl <- fread(sprintf("zcat %s","/home/peter/prostate_cancer/colocalization/COLOC/GTEx_Analysis_v8_eQTL/" %&% tissues[tissue] %&% ".v8.signif_variant_gene_pairs.txt.gz")) #Unzips and reads in the gene-SNP pair eQTL file
  meQTL_for_COLOC <- meqtl %>% dplyr::select(gene_id, variant_id, maf, pval_nominal, slope, slope_se) #The file originally had 12 columns.  This subsets to COLOC input to six columns
  #meQTL_for_COLOC <- subset(meQTL_for_COLOC, meQTL_for_COLOC$gene_id == "ENSG00000123179.13") #This subsets the data town further to out gene of interest
  meQTL_for_COLOC <- subset(meQTL_for_COLOC, variant_id %in% GWAS_SNPs) #This cuts the eQTL data to only those in the GWAS file
  meQTL_for_COLOC$n_samples <- tissues_sample_size[tissue] #This generates the tissue sample size as a column in the eQTL data
  colnames(meQTL_for_COLOC) <- c("gene_id", "variant_id", "maf", "pval_nominal", "slope", "slope_se", "n_samples")
  meQTL_for_COLOC <- meQTL_for_COLOC[complete.cases(meQTL_for_COLOC),] # This selects only complete cases for the eqtl file
  eQTL_write <- rbind(eQTL_write, meQTL_for_COLOC) #This sets the eQTL data to an object that can be written
  for(chr in chrs){
    print("Reading chromosome " %&% chr)
    GWAS_result <- fread("/home/peter/prostate_cancer/colocalization/COLOC/chr_files/chr" %&% chr %&% ".txt", header= F) #Reads in a single chromosome GWAS
    colnames(GWAS_result)<-c("CHR", "SNP", "BP", "A1", "TEST", "NMISS", "OR", "STAT", "P", "BETA", "SE", "MAF", "SNP_ID")# Assigns column names to the chromosome GWAS file
    GWAS_for_COLOC <- GWAS_result %>% dplyr::select(SNP_ID, BETA, SE, MAF) #The original file had a lot more columns, as shown above.  This cuts the data to the four columns we care about
    GWAS_for_COLOC$sample_size <- sample_size #This adds a sample size column to the data object
    colnames(GWAS_for_COLOC) <- c("panel_variant_id", "effect_size", "standard_error", "frequency", "sample_size") #This renames the columns
    GWAS_for_COLOC <- GWAS_for_COLOC[complete.cases(GWAS_for_COLOC),] #COLOC does not like missing values
    GWAS_write <- rbind(GWAS_write, GWAS_for_COLOC) #This sets the GWAS data to an object that can be written
  }
  snps_in_both <- intersect(GWAS_write$panel_variant_id, eQTL_write$variant_id) #is there a better way to do this? Probably. Do I feel like figuring it out? Nah.
  snps_in_all <- intersect(snps_in_both, GWAS_SNPs) #only keep overlapping SNPs
  GWAS_write <- subset(GWAS_write, panel_variant_id %in% snps_in_all)
  eQTL_write <- subset(eQTL_write, variant_id %in% snps_in_all)
  eQTL_write <- eQTL_write[order(eQTL_write$gene_id),] #results are weird when not ordered
  
  fwrite(eQTL_write, "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_eQTL_" %&% tissues[tissue] %&% ".txt", quote = F, sep = "\t", na = "NA", row.names = F, col.names = T)
  gzip("/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_eQTL_" %&% tissues[tissue] %&% ".txt", destname = "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/" %&% "AA_pro_can" %&% "_eQTL_" %&% tissues[tissue] %&% ".txt.gz", overwrite = T) #script may only take .gz values so can't hurt to be too careful
  fwrite(GWAS_write, "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_GWAS_" %&% tissues[tissue] %&% ".txt", row.names = F, col.names = T, sep = "\t", quote = F, na = "NA")
  gzip("/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_GWAS_" %&% tissues[tissue] %&% ".txt", destname = "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/" %&% "AA_pro_can" %&% "_GWAS_" %&% tissues[tissue] %&% ".txt.gz", overwrite = T)
  print("Completed with " %&% tissues[tissue] %&% ", for " %&% "AA_pro_can .")
  print("Completed making input for COLOC.")
}




```
I think it is probably best to break the script into three pieces: Tissue, GWAS, and Writing:

### Tissue
Here, we read in gene-SNP eQTL data from GTEx.  We will be fcusing mostly on the gene, SNP, MAF, p-value, beta, and standard error.  Once we select all of those, we take only the SNPs found in our GWAS.  We then assign the sample count for each tissue.

### GWAS
We read in one chromosome at a time for the GWAS files.  The positions are in hg38 build, but the names in the `SNP` column are still named as rsIDs. The `SNP_ID` column has the SNPs named as they are in GTEx v8.  In this data, we want the hg38 name, beta, standard error, and MAF.  We assign a sample size to each GWAS file (They are all the same).  After that, we find the SNPs that overlap at the end between the GWAS and eQTL file again.

### Writing
Lastly, we write and zip the files to the director created in the initial set up.
```{r}
snps_in_both <- intersect(GWAS_for_COLOC$panel_variant_id, eQTL_write$variant_id) #is there a better way to do this? Probably. Do I feel like figuring it out? Nah.
snps_in_all <- intersect(snps_in_both, GWAS_SNPs) #only keep overlapping SNPs
GWAS_write <- subset(GWAS_for_COLOC, panel_variant_id %in% snps_in_all)
eQTL_write <- subset(eQTL_write, variant_id %in% snps_in_all)
eQTL_write <- eQTL_write[order(eQTL_write$gene_id),] #results are weird when not ordered
  
fwrite(eQTL_write, "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_eQTL_" %&% tissues[tissue] %&% ".txt", quote = F, sep = "\t", na = "NA", row.names = F, col.names = T)
gzip("/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_eQTL_" %&% tissues[tissue] %&% ".txt", destname = "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/" %&% AA_pro_can %&% "_eQTL_" %&% tissues[tissue] %&% ".txt.gz", overwrite = T) #script may only take .gz values so can't hurt to be too careful
fwrite(GWAS_write, "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_GWAS_" %&% tissues[tissue] %&% ".txt", row.names = F, col.names = T, sep = "\t", quote = F, na = "NA")
gzip("/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/AA_pro_can" %&% "_GWAS_" %&% tissues[tissue] %&% ".txt", destname = "/home/peter/prostate_cancer/colocalization/COLOC/COLOC_input/" %&% AA_pro_can %&% "_GWAS_" %&% tissues[tissue] %&% ".txt.gz", overwrite = T)
print("Completed with " %&% tissues[tissue] %&% ", for " %&% "AA_pro_can .")
print("Completed making input for COLOC.")
```



# Running COLOC
Now we literally, just plug in the files we just made into COLOC.  The script `run_COLOC_wrapper.py` does this for us.  We just need to edit our GWAS and eQTL files. Similar to the script above, we also enter sample size and phenotype.